Pythran: Bringing OpenMP to a Python Subset

### 2013-02-11 doc/papers/europar2013/paper.tex
Correct EuroPar 2013 article up to section 2 inclusive.
Be strict with the names of other projects!

Copperhead~\cite{copperhead2011} is a functional, data parallel language
embedded in Python. It uses n-uplets, numpy arrays and lists as its core data
structure and prohibits usage of many control-flow operators such as loops,
enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
intrinsics to exhibit parallelism. But it can be efficiently compiled to either
CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.\
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are JIT-compiled to native code.

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shedskin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
turn regular Python codes into optimized ones, but they do not offer support for
fine-grained parallelism beyond what the standard library proposes.
%~\footnote{Shedskin is compatible with several Python modules that provide
%coarse-grained, fork-based parallelism.}

@@

This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
with a Python subset called Pythran to bring fine-grain parallelism to Python.
While being backward compatible with both the Python language, and the
sequential algorithm. It means that:
%
\begin{enumerate}
    \item Any Pythran code can be run (sequentially), with no module dependency
        or code change, by any Python interpreter.
    \item Parallelism is explicit and incrementally added to the original code
        through directives.
\end{enumerate}

@@

A critical point of the design of the \texttt{pythonic++} runtime is memory
management, the very same aspect that led to the use of a GIL in CPython.
Memory management is implemented in Shedskin through the general-purpose Boehm
garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
objects inside parallel regions, thus making memory management explicit for
these parts.

@@

The traditional solution is to use a native type that exposes a Python interface
and has a constant translation cost. Numpy's \texttt{ndarray} is a typical
implementation of this concept and is the keystone of the Scipy and Numpy
packages. Basically, a numpy \texttt{ndarray} is a raw C pointer that is exposed
at the Python level. As a tool for scientific computing, Pythran supports such a
structure.

>>>>

Copperhead~\cite{copperhead2011} is a functional, data parallel language
embedded in Python. It uses $n$-uplets, NumPy arrays and lists as its core data
structure and prohibits usage of many control-flow operators such as loops,
enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
intrinsics to exhibit parallelism. But it can be efficiently compiled to either
CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.\
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are JIT-compiled to native code.

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
turn regular Python codes into optimized ones, but they do not offer support for
fine-grained parallelism beyond what the standard library proposes.
%~\footnote{Shedskin is compatible with several Python modules that provide
%coarse-grained, fork-based parallelism.}

@@

This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
with a Python subset called Pythran to bring fine-grain parallelism to Python,
while being backward compatible with both the Python language and the
sequential algorithm. It means that:
%
\begin{enumerate}
    \item any Pythran code can be run (sequentially), with no module dependency
        or code change, by any Python interpreter;
    \item parallelism is explicit and incrementally added to the original code
        through directives.
\end{enumerate}

@@

A critical point of the design of the \texttt{pythonic++} runtime is memory
management, the very same aspect that led to the use of a GIL in CPython.
Memory management is implemented in Shed Skin through the general-purpose Boehm
garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
objects inside parallel regions, thus making memory management explicit for
these parts.

@@

The traditional solution is to use a native type that exposes a Python interface
and has a constant translation cost. NumPy's \texttt{ndarray} is a typical
implementation of this concept and is the keystone of the Scipy and NumPy
packages. Basically, a NumPy \texttt{ndarray} is a raw C pointer that is exposed
at the Python level. As a tool for scientific computing, Pythran supports such a
structure.
